Whitepaper: Contract Modules

### 2021-05-14 papers/contract-modules/modules.tex
abstract

\lstset{
  language=caml,
  morekeywords={contract,entrypoint,module,hash,paid,raises,sig,val},
  % predefined types
  classoffset=1,
  morekeywords={nat,int,unit,string,pukh,status,tz},
  keywordstyle=\color{blue},
  classoffset=2,
  morekeywords={Left,Right},
  keywordstyle=\color{magenta},
  backgroundcolor=\color{brown!20},
  captionpos=b,
  literate={->}{{$\to$}}1
}

@@

\section{Motivation}
\label{sec:motivation}

Consider this Michelson code fragment:
\begin{lstlisting}[language=michelson,caption={Michelson code example},label={lst:code-example}]
parameter (or (int %bid_for_auction)
              (unit %close_auction));
storage address;
code { DUP; CAR; 
       IF_LEFT {
       # %bid_for_auction
        DUP; PUSH int 10000; SUB; LE;
        IF { FAILWITH; } {} # fails with argument value
        # ...
        { PUSH string "closed"; FAILWITH; }
        # ...
        { PUSH string "bid too low"; FAILWITH; }
        # ...
        { PUSH int 42; FAILWITH; }
        # ...
       }
       {
       # %close_auction
         DROP; CDR; SENDER; EQ;
         IF {} { PUSH string "not owner"; FAILWITH; }
         # ...
       }
     }
\end{lstlisting}

Here is a \textbf{contract module} that describes / specifies the above Michelson code.
\begin{lstlisting}[caption={Contract module example},label={lst:contract-module-example}]
contract type MyContract = sig
  paid entrypoint %bid_for_auction (v : int) 
  raises "closed"                 (** auction closed*)
       | "bid too low" as too_low (** bid received is too low *)
       | 42 as the_answer         (** failwith received argument 42 *)
       | (_ : int) as failedwith  (** failwith received an integer arg *)

  entrypoint %close_auction (_ : unit)
  raises "not owner"         (** caller cannot close the auction *)
end
\end{lstlisting}

@@

Each \lstinline/entrypoint/ is mapped to a function. The variables
used in the pattern may be a wildcard \lstinline/_/ or an OCaml identifier, which
is used in generating the implementation. For \lstinline/_/ a standard
name will be generated.

A \lstinline/paid/ entrypoint requires the caller to specify an amount of tokens to
accompany the contract invocation. Accordingly, the generated function has an extra
argument. Otherwise, the entrypoint is unpaid.

The \lstinline/raises/ clauses specify values that the contract might
pass to the \texttt{FAILWITH} instruction. The \lstinline/as/ clause
maps the matched value to an OCaml identifier, which is used to
generate the corresponding case in the error type. There is a standard
translation (to be specified) that maps a value without an
\lstinline/as/ clause to an identifier.

Alternatively and equivalently, the contracts can be specified using
patterns on the default entrypoint.

\begin{lstlisting}[caption={Alternative syntax for entrypoints},label={lst:alternative-syntax}]
  paid entrypoint bid_for_auction (Left (_ : int))
...
  entrypoint close_auction (Right (_ : unit))
\end{lstlisting}

@@

\begin{lstlisting}[caption={Generated signature},label={lst:generated-signature}]
module type MyContract =
sig
  type bid_for_auction_errors = 
     | closed       (** auction closed*)
     | too_low      (** bid received is too low *)
     | the_answer   (** failwith received argument 42 *)
     | failedwith of int (** failwith received an integer arg *)

  val bid_for_auction
    : Tezos.pukh -> ?fee:Tezos.tz -> amount:Tezos.tz
    -> int
    -> (Tezos.status, bid_for_auction_errors) Tezos.monad

  type close_auction_errors = 
     | not_owner    (** caller cannot close the auction *)

  val close_auction
    : Tezos.pukh -> ?fee:Tezos.tz -> ?amount:Tezos.tz
    -> unit
    -> (Tezos.status, close_auction_errors) Tezos.monad
end
\end{lstlisting}

@@

  Q (extension): Would it make sense to specify something about fees in the
  contract module? For example, we might specify a fee schedule which
  raises the fee until the call is accepted.
\item The type \lstinline/Tezos.monad/ is defined by the underlying
  API. The untyped API uses a type \lstinline/Answer.t/. The exact
  relationship is TBD.
\item Error reporting assumes that the monad returns a sum type. But
  things could be handled differently (for the
  \lstinline/close_auction/entrypoint): by an error continuation of
  type \lstinline/close_auction_errors -> Answer.t/ or by
  raising an exception that encapsulates the error type:
\begin{lstlisting}
exception Close_Auction_Exn of close_auction_errors
\end{lstlisting}
\end{itemize}

@@

type bid_for_auction_errors = 
   | closed       (** auction closed*)
   | too_low      (** bid received is too low *)
   | the_answer   (** failwith received argument 42 *)
   | failedwith of int (** failwith received an integer arg *)

let big_for_auction_errors_map =
  let open Base in
  Base.Map.of_alist_exn (module String)
    [
      (failwith_regex ^ "closed", closed);
      (failwith_regex ^ "bid too low", too_low);
      (failwith_regex ^ "42", the_answer);
      (failwith_regex ^ int_regex, failedwith (0)); 
      (* badness 10000: actual implementation must obtain the int value *)
    ]
  (* inadequate in several ways;
   * should just generate code instead of using a map*)

let bid_for_auction destination ?(fee=100.0) amount v =
  ...
  begin
    SyncAPIV0.call_contract
      amount
      destination
      contract_addr
      ~(serialize Tint v)
      fee
    >>=? fun oph ->
    wait_for_inclusion oph (* busy waiting *)
    >>=? fun _ ->
    return 0
  end
  >>= function
  | Ok _ -> ...
  | Error (Rejection Michelson_runtime_error s) -> 
    (*match s against expected errors from ...errors_map *)
    (*throw exception if no match *)
    ...

type close_auction_errors = 
   | not_owner    (** caller cannot close the auction *)

let close_auction destination ?(fee=50.0) ?(amount=0.0) v =
  ...
  SyncAPIV0.call_contract
    amount
    destination
    contract_addr
    ~(serialize Tunit v)
    fee
\end{lstlisting}
This code example is incomplete and likely incorrect. It contains
fragments from an example implemented on top of the synchronous API.
The call to \lstinline/serialize/ refers to a proposal to improve on
the typing properties of the synchronous API.

@@

The recommended technology for the type checker is symbolic
execution. In this framework, it would also 
be possible to specify further invariants about the inputs and place them into the contract
module. Essentially, every formula that an SMT solver can deal with
should be fine, although a first version of the symbolic executor
could work without using SMT.

@@

 The stack is initialized with a symbolic value corresponding to a pair
of the storage type and the parameter type. The latter is concretized
according to the path to the specified entrypoint. That is, to
investigate the entrypoint \lstinline/bid_for_auction/ in
Listing~\ref{lst:code-example}, symbolic execution would start with a
one-element stack that contains
\begin{lstlisting}[language=michelson]
(PAIR (LEFT x1:int) x2:address)
\end{lstlisting}
without any constraints on \texttt{x1} or \texttt{x2}.

>>>>

\lstset{
  language=caml,
  morekeywords={requires,sender,amount,balance,storage,contract,entrypoint,module,hash,paid,raises,sig,val},
  % predefined types
  classoffset=1,
  morekeywords={nat,int,unit,string,pukh,status,tz},
  keywordstyle=\color{blue},
  classoffset=2,
  morekeywords={Pair,Left,Right},
  keywordstyle=\color{magenta},
  backgroundcolor=\color{brown!20},
  captionpos=b,
  literate={->}{{$\to$}}1
}

@@

\section{A simple example}
\label{sec:motivation}

Consider this Michelson code fragment for implementing an auction contract:
\begin{lstlisting}[language=michelson,caption={Michelson code example},label={lst:code-example}]
parameter (or (unit %close)
              (unit %bid));
storage (pair bool (pair address address));
code { DUP; CAR; 
       IF_LEFT {
       # %close
         DROP; CDR; CAR; SENDER; EQ;
         IF {} { PUSH string "not owner"; FAILWITH; }
         # ...
       }
       {
       # %bid
        DROP; AMOUNT; PUSH mutez 10000; COMPARE; LE;
        IF { FAILWITH; } {} # fails with argument value
        # ...
        { PUSH string "closed"; FAILWITH; }
        # ...
        { PUSH string "bid too low"; FAILWITH; }
        # ...
        { PUSH int 42; FAILWITH; }
        # ...
       }
     }
\end{lstlisting}

Here is a simple \textbf{contract module} that describes / specifies
the above Michelson code. 
\begin{lstlisting}[caption={Contract module example},label={lst:contract-module-example}]
contract type MyContract = sig
  paid entrypoint %bid () 
  raises "closed"                 (** auction closed*)
       | "bid too low" as too_low (** bid received is too low *)
       | 42 as the_answer         (** failwith received argument 42 *)
       | (_ : int) as failedwith  (** failwith received an integer arg *)

  entrypoint %close ()
  raises "not owner"         (** caller cannot close the auction *)
end
\end{lstlisting}

@@

Each \lstinline/entrypoint/ is mapped to a function. The entrypoint's
parameter is described by a pattern.  Variables (OCaml identifier) or
wildcards \lstinline/_/ in the pattern will be used as formal
parameters in generating the implementation. For \lstinline/_/ a
standard name will be generated. Each variable or wildcard needs a
type annotation. Constructors in the pattern will be
omitted from the interface, but automatically inserted in the contract
invocation.

A \lstinline/paid/ entrypoint requires the caller to specify an amount of tokens to
accompany the contract invocation. Accordingly, the generated function
has a mandatory argument to specify this amount. Otherwise, the entrypoint is unpaid.

The \lstinline/raises/ clause specifies values that the contract might
pass to the \texttt{FAILWITH} instruction. The translation generates
an error type from the clause that describes all possible
failures. The \lstinline/as/ clause 
maps the matched value to an OCaml identifier, which is used to
generate the corresponding case in the error type. There is a standard
translation (to be specified) that maps a value without an
\lstinline/as/ clause to an identifier.

The \lstinline/raises/ clause must be exhaustive. Any failure arising
during execution of the contract must be captured by one of the
specified cases. If in doubt, include a catchall default case.

Alternatively and equivalently, contracts can be specified using
patterns on the default entrypoint.

\begin{lstlisting}[caption={Alternative syntax for entrypoints},label={lst:alternative-syntax}]
  paid entrypoint bid (Right ())
...
  entrypoint close (Left ())
\end{lstlisting}

The specification does not have to be exhaustive. The contract may
have further entrypoints that are not described by the module.

\clearpage{}
\section{A safer example}
\label{sec:safe-example}

Alternatively, one may specify preconditions that are checked before
the contract is deployed to the blockchain.

\begin{lstlisting}[caption={Safer contract module},label={lst:safer-contract-module}]
contract type SaferContract = sig
  storage (Pair (bidding : bool) 
                (Pair (owner : address) 
                      (highest_bidder : address)))

  entrypoint %close ()
  requires (sender = owner) raises "not owner"

  paid entrypoint %bid () 
  requires bidding raises "closed" (** auction closed*)
  requires (amount > balance)
  requires (amount > 10000)
  raises "bid too low" as too_low (** bid received is too low *)
       | 42 as the_answer         (** failwith received argument 42 *)
end
\end{lstlisting}
The \lstinline/storage/ clause defines a typed pattern to match the
current storage of the contract. Moreover, it generates a suitably
typed getter function (also named \lstinline/storage/) 
to access the current storage from OCaml.

The \lstinline/%close/ entrypoint comes with a \lstinline/requires/
clause that specifies a precondition for invoking the contract. This
precondition is checked in the OCaml interface \emph{before} invoking
the contract. The entrypoint does not have a \lstinline/raises/
clause anymore, which means that this precondition implies that no
failures can occur.

\textbf{For this reasoning to be true, we need to know that the owner
  component of the storage is constant through all contract executions.}

For the \lstinline/%bid/ entrypoint, the \lstinline/bidding/ component
must be true. However, we cannot guarantee that there is nobody
closing the contract between accessing the storage and the
actual contract invocvation. Hence, the corresponding
\lstinline/raises/ clause must remain in place.

\textbf{To safely reject when \lstinline/bidding/ is false, we need to know
  that there is no contract execution that can make it true, again.}

Analogous reasoning applies to the condition
\lstinline/amount > balance/. Here, we also need to know that the
balance cannot decrease, which is non-trivial. However, we can accurately check that
\lstinline/amount > 10000/ before the contract invocation, which rules
out the unspecific integer failure.

The interface of the generated functions remains the same, but some of
the errors are generated without involving the blockchain.

\section{Generated code}
\label{sec:generated-code}

@@

\begin{lstlisting}[caption={Generated signature},label={lst:generated-signature}]
module type MyContract =
sig
  type bid_errors = 
     | closed       (** auction closed*)
     | too_low      (** bid received is too low *)
     | the_answer   (** failwith received argument 42 *)
     | failedwith of int (** failwith received an integer arg *)

  val bid
    : Tezos.pukh -> ?fee:Tezos.tz -> amount:Tezos.tz
    -> (Tezos.status, bid_errors) Tezos.monad

  type close_errors = 
     | not_owner    (** caller cannot close the auction *)

  val close
    : Tezos.pukh -> ?fee:Tezos.tz -> ?amount:Tezos.tz
    -> (Tezos.status, close_errors) Tezos.monad
end
\end{lstlisting}

@@

  Q (extension): Would it make sense to specify something about fees in the
  contract module? For example, we might specify a fee schedule which
  raises the fee until the call is accepted.
\item The type \lstinline/Tezos.monad/ is defined by the underlying
  API. The untyped API uses a type \lstinline/Answer.t/. The exact
  relationship is TBD.
\item In this example, there are no parameters that are passed to the
  entrypoints because their type is \lstinline/Unit/.
\item Error reporting assumes that the monad returns a sum type. But
  things could be handled differently (for the
  \lstinline/close/ entrypoint): by an error continuation of
  type \lstinline/close_errors -> Answer.t/ or by
  raising an exception that encapsulates the error type:
\begin{lstlisting}
exception Close_Exn of close_errors
\end{lstlisting}
\end{itemize}

@@

type bid_errors = 
   | closed       (** auction closed*)
   | too_low      (** bid received is too low *)
   | the_answer   (** failwith received argument 42 *)
   | failedwith of int (** failwith received an integer arg *)

let big_errors_map =
  let open Base in
  Base.Map.of_alist_exn (module String)
    [
      (failwith_regex ^ "closed", closed);
      (failwith_regex ^ "bid too low", too_low);
      (failwith_regex ^ "42", the_answer);
      (failwith_regex ^ int_regex, failedwith (0)); 
      (* badness 10000: actual implementation must obtain the int value *)
    ]
  (* inadequate in several ways;
   * should just generate code instead of using a map*)

let bid destination ?(fee=100.0) amount v =
  ...
  begin
    SyncAPIV0.call_contract
      amount
      destination
      contract_addr
      ~(serialize Tint v)
      fee
    >>=? fun oph ->
    wait_for_inclusion oph (* busy waiting *)
    >>=? fun _ ->
    return 0
  end
  >>= function
  | Ok _ -> ...
  | Error (Rejection Michelson_runtime_error s) -> 
    (*match s against expected errors from ...errors_map *)
    (*throw exception if no match *)
    ...

type close_errors = 
   | not_owner    (** caller cannot close the auction *)

let close destination ?(fee=50.0) ?(amount=0.0) v =
  ...
  SyncAPIV0.call_contract
    amount
    destination
    contract_addr
    ~(serialize Tunit v)
    fee
\end{lstlisting}
This code example is incomplete and likely incorrect. It contains
fragments from an example implemented on top of the synchronous API.
The call to \lstinline/serialize/ refers to a proposal to improve on
the typing properties of the synchronous API.

@@

The recommended technology for the type checker is symbolic
execution. In this framework, it would also 
be possible to specify further invariants (like the
\lstinline/require/ clauses) about the inputs and place
them into the contract module. Essentially, every formula that an SMT
solver can deal with should be fine, although a first version of the
symbolic executor could work without using SMT.

@@

 The stack is initialized with a symbolic value corresponding to a pair
of the storage type and the parameter type. The latter is concretized
according to the path to the specified entrypoint. That is, to
investigate the entrypoint \lstinline/bid/ in
Listing~\ref{lst:code-example}, symbolic execution would start with a
one-element stack that contains
\begin{lstlisting}[language=michelson]
(PAIR (LEFT x1:int) x2:address)
\end{lstlisting}
without any constraints on \texttt{x1} or \texttt{x2}.
